High-Dimensional Regularized Discriminant Analysis

### 2016-01-20 reports/paper-hdrda.Rnw
Added some DMY edits.

\cite{Friedman:1989tm} proposed the popular regularized discriminant analysis
(RDA) classifier that utilizes a biased covariance-matrix estimator that
partially pools the sample covariance matrices from linear and quadratic
disriminant analysis and shrinks the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure using a
grid search. We formulate a new RDA classifier for the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to a classifier in a subspace having a much lower dimension. As a
result, the utilization of the dimension-reduction step yields a substantial
reduction in computation during model selection. Also, our parameterization
offers interpretability previously lacking with the RDA classifier. We
demonstrate that our proposed classifier is often superior to several recently
proposed sparse and regularized classifiers in terms of classification accuracy
with three artificial and six real high-dimensional data sets. Finally, we
provide an implementation of our proposed classifier in the {\tt sparsediscrim}
R package, which is available on CRAN.

@@

In Figure \ref{fig:sim-results}, we observe each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the for the Pang, Tong, and
Witten classifiers. For smaller dimensions, the random-forest and \emph{HDRDA}
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both \emph{HDRDA} classifiers outperformed all other classifiers, suggesting
that their inherent dimension reduction better captured the classificatory
information in the small samples even in the presence of outliers.

@@

The Pang, Tong, and Witten methods yielded practically the same and consistently
the worst error rates when outliers were present with $\epsilon > 0$, suggesting
that these classifiers were sensitive to outliers. Notice, for example, that
when $p=400$, the error rates of the Pang, Tong, and Witten classifiers
increased dramatically from approximately 19\% when no outliers were present to
approximately 43\% when $\epsilon=0.05$. The sharp increase in average error
rates for these three classifiers continued as $\epsilon$ increased. Guo's
method always outperformed Pang, Witten, and Tong, but once outliers were
present, the Guo classifier's average error rate was competitive with neither
the \emph{HDRDA} classifiers nor the random forest.

@@

For $\epsilon > 0$, we observed a different behavior in classification
performance. First, the Pang, Tong, and Witten methods along with the random
forest increased in average error rate as $p$ increased. Contrarily, the
performance of the \emph{HDRDA} and Guo classifiers was hardly affected by $p$.
Also, as discussed above, the \emph{HDRDA} classifiers were superior to all
other classifiers for larger values of $p$ with only the random forest
outperforming them for smaller feature dimensions.

@@

We compared the \emph{HDRDA} classifer to the five competing classifiers on six
benchmark gene-expression microarray data sets. We evaluated the classification
accuracy of each classifier by randomly partitioning the data set under
consideration such that $2/3$ of the observations were allocated as training
data and the remaining $1/3$ of the observations were allocated as a test data
set. To expedite the computational runtime, we reduced the training data to the
top 1000 variables by employing the variable-selection method proposed by
\cite{Dudoit:2002ev}. We then reduced the test data set to the same 1000
variables. After training each classifier on the training data, we classified
the test data sets and computed the proportion of mislabeled test observations
to estimate the classification error rate for each classifier. Repeating this
process 100 times, we computed the average of the error-rate estimates for each
classifier. Here, we provide a concise description of each high-dimensional data
set examined in our classification study.

@@

The average error-rate estimates for the Pang, Tong, and Witten classifiers were
comparable across all data sets. Furthermore, the average error rates for the
Pang and Tong classifiers were approximately equal for all data sets except for
the Chiaretti dataset. This result suggests that the mean and variance
estimators used in lieu of the MLEs provided little improvement to
classification accuracies. However, we investigated the Pang classifier's poor
performance on the Chiaretti data set and determined that its variance estimator
exhibited numerical instability. The classifier's denominator was approximately
zero for both classes and lead to the poor classification performance.

@@

Finally, the Pang, Tong, and Witten classifiers consistently yielded the largest
error rates across all data sets. Given that the standard deviations were
relatively large, we hesitate to generalize claims regarding the ranking of
these three classifiers in terms of error rate. However, the classifiers' error
rates and their variability across multiple random partitions of each data set
were large enough to rethink applying them in practice.

>>>>

\cite{Friedman:1989tm} proposed the popular regularized discriminant analysis
(RDA) classifier that utilizes a biased covariance-matrix estimator that
partially pools the sample covariance matrices from linear and quadratic
disriminant analysis and shrinks the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure using a
grid search. We formulate a new RDA-based classifier for the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to a classifier in a subspace having a much lower dimension. As a
result, the utilization of the dimension-reduction step yields a substantial
reduction in computation during model selection. Also, our parameterization
offers interpretability that was previously lacking with the RDA classifier. We
demonstrate that our proposed classifier is often superior to several recently
proposed sparse and regularized classifiers in terms of classification accuracy
with three artificial and six real high-dimensional data sets. Finally, we
provide an implementation of our proposed classifier in the {\tt sparsediscrim}
R package, which is available on CRAN.

@@

In Figure \ref{fig:sim-results}, we observe each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the for the Pang, Tong, and
Witten classifiers. For smaller dimensions, the random-forest and \emph{HDRDA}
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both \emph{HDRDA} classifiers outperformed all other classifiers, suggesting
that their inherent dimension reduction better captured the classificatory
information in the small training samples, even in the presence of outliers.

@@

The Pang, Tong, and Witten methods yielded practically the same and consistently
the worst error rates when outliers were present with $\epsilon > 0$, suggesting
that these classifiers were sensitive to outliers. Notice, for example, that
when $p=400$, the error rates of the Pang, Tong, and Witten classifiers
increased dramatically from approximately 19\% when no outliers were present to
approximately 43\% when $\epsilon=0.05$. The sharp increase in average error
rates for these three classifiers continued as $\epsilon$ increased. Guo's
method always outperformed Pang, Witten, and Tong, but once outliers were
present, the Guo classifier's average error rate was not competitive with
the \emph{HDRDA} classifiers or the random-forest classifier.

@@

For $\epsilon > 0$, we observed a different behavior in classification
performance. First, the Pang, Tong, and Witten methods, along with the
random-forest method increased in average error rate as $p$
increased. Contrarily, the performance of the \emph{HDRDA} and Guo classifiers
was hardly affected by $p$.  Also, as discussed above, the \emph{HDRDA}
classifiers were superior to all other classifiers for large values of $p$ with
only the random-forest classifier outperforming them in smaller
feature-dimension cases.

@@

We compared the \emph{HDRDA} classifer to the five competing classifiers on six
benchmark gene-expression microarray data sets. We evaluated the classification
accuracy of each classifier by randomly partitioning the data set under
consideration such that $2/3$ of the observations were allocated as training
data and the remaining $1/3$ of the observations were allocated as a test data
set. To expedite the computational runtime, we reduced the training data to the
top 1000 variables by employing the variable-selection method proposed by
\cite{Dudoit:2002ev}. We then reduced the test data set to the same 1000
variables. After training each classifier on the training data, we classified
the test data sets and computed the proportion of mislabeled test observations
to estimate the classification error rate for each classifier. Repeating this
process 100 times, we computed the average of the error-rate estimates for each
classifier. We next provide a concise description of each high-dimensional data
set examined in our classification study.

@@

The average error-rate estimates for the Pang, Tong, and Witten classifiers were
comparable across all six data sets. Furthermore, the average error rates for the
Pang and Tong classifiers were approximately equal for all data sets except for
the Chiaretti dataset. This result suggests that the mean and variance
estimators used in lieu of the MLEs provided little improvement to
classification accuracies. However, we investigated the Pang classifier's poor
performance on the Chiaretti data set and determined that its variance estimator
exhibited numerical instability. The classifier's denominator was approximately
zero for both classes and lead to the poor classification performance.

@@

Finally, the Pang, Tong, and Witten classifiers consistently yielded the largest
average error rates across the six data sets. Given that the standard deviations
were relatively large, we hesitate to generalize claims regarding the ranking of
these three classifiers in terms of the average error rate. However, the
classifiers' error rates and their variability across multiple random partitions
of each data set were large enough that we might question their benefit when
applied to real data.
